---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 3 - Due date 02/01/24"
author: "Jaimie Wargo"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A03_Sp24.Rmd"). Then change "Student Name" on line 4 with your name.

Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai.

## Questions

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only.

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

```{r Packages}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
library(tidyverse)
library(here)
library(cowplot)
library(readxl)
```

## Trend Component

### Q1

For each time series, i.e., Renewable Energy Production and Hydroelectric Consumption create three plots: one with time series, one with the ACF and with the PACF. You may use the same code from A2, but I want all the three plots side by side as in a grid. (Hint: use function `plot_grid()` from the `cowplot` package)

```{r Data Import}
#Importing data set
raw_data <- read_excel(path=here('Data', 
          'Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx'),
          skip = 12, sheet="Monthly Data",col_names=FALSE) 

#Extract the column names from row 11
read_col_names <- read_excel(path=here('Data', 
          'Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx'),
            skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 

colnames(raw_data) <- read_col_names

#Data frame
energy_df <- data.frame(raw_data[,c('Month', 'Total Biomass Energy Consumption', 
                         'Total Renewable Energy Consumption', 
                         'Hydroelectric Power Consumption')])

nospace_colnames <- c('date', 'biomass', 'total_renew', 'hydro')
colnames(energy_df) <- nospace_colnames

#Time series structure
energy_ts <- ts(energy_df[,2:4], start=c(1973, 1), frequency = 12)
```

```{r Biomass Plots}

bm_tsplot <- ggplot(data=energy_df, aes(x=date, y=biomass))+
  geom_line(color='darkseagreen')+
  labs(x='Date', y='Biomass (Trillion Btu)', 
       title='Biomass Energy over time', subtitle = '1973-2023')

bm_acfplot <- autoplot(Acf(energy_ts[,1], lag.max=40, plot=F), main="Biomass ACF")
                       
bm_pacfplot <- autoplot(Pacf(energy_ts[,1], lag.max=40, plot=F), main="Biomass PACF")

plot_grid(bm_tsplot, bm_acfplot, bm_pacfplot)
```

```{r Hydropower Plots}

hp_tsplot <- ggplot(data=energy_df, aes(x=date, y=hydro))+
  geom_line(color='steelblue4')+
  labs(x='Date', y='Hydropower (Trillion Btu)', 
       title='Hydropower over time', subtitle = '1973-2023')

hp_acfplot <- autoplot(Acf(energy_ts[,3], lag.max=40, plot=F), main="Hydropower ACF")
                       
hp_pacfplot <- autoplot(Pacf(energy_ts[,3], lag.max=40, plot=F), main="Hydropower PACF")

plot_grid(hp_tsplot, hp_acfplot, hp_pacfplot)
```

```{r Total Renewable Plots}

tre_tsplot <- ggplot(data=energy_df, aes(x=date, y=total_renew))+
  geom_line(color='plum4')+
  labs(x='Date', y='Renewable Energy (Trillion Btu)', 
       title='Renewable Energy over time', subtitle = '1973-2023')

tre_acfplot <- autoplot(Acf(energy_ts[,2], lag.max=40, plot=F),main="Renewable Energy ACF")
                       
tre_pacfplot <- autoplot(Pacf(energy_ts[,2], lag.max=40,plot=F),main="Renewable Energy PACF")

plot_grid(tre_tsplot, tre_acfplot, tre_pacfplot)
```

### Q2

From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

Answer: The biomass and total renewable energy plots both seem to have an upwards trend, whereas hydroelectric appears constant over the course of the observations.

### Q3

Use the *lm()* function to fit a linear trend to the three time series (I am going to just use total renewable and hydro as indicated in lecture 2/1). Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r}

nobs <- nrow(energy_df)
t <- 1:nobs

renewable_trend <- lm(energy_df[,3]~t)
summary(renewable_trend)

tre_beta0 <- renewable_trend$coefficients[1]
tre_beta1 <- renewable_trend$coefficients[2]
```

The R2 of 0.8131 indicates that 81.31% of variation can be explained by the trend. The overall slope is estimated at 0.685, indicating an increasing trend, and the intercept of 184 indicates the starting value as modeled linearly, starting at 184 trillion Btu of renewable energy in 1973.

```{r}
hydro_trend <- lm(energy_df[,4]~t)
summary(hydro_trend)

hp_beta0 <- hydro_trend$coefficients[1]
hp_beta1 <- hydro_trend$coefficients[2]
```

The R2 of 0.013 indicates that 1.3% of variation can be explained by the trend. The overall slope is estimated at -0.00985, indicating a fairly flat trend (or lack thereof), and the intercept of 82.73 indicates the starting value as modeled linearly, starting at 82.73 trillion Btu of hydroelectric energy in 1973.

### Q4

Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

```{r}
tre_detrend <- energy_df[,3] - (tre_beta0+tre_beta1*t)

tre_df <- data.frame('date'=energy_df$date,
                         'observed'=energy_df[,3],
                         'detrend'=tre_detrend)

ggplot(tre_df, aes(x=date))+
  geom_line(aes(y=observed), color='plum4')+
  geom_line(aes(y=detrend), color='purple4')
```

```{r}
hp_detrend <- energy_df[,4] - (hp_beta0+hp_beta1*t)

hp_df <- data.frame('date'=energy_df$date,
                         'observed'=energy_df[,4],
                         'detrend'=hp_detrend)

ggplot(hp_df, aes(x=date))+
  geom_line(aes(y=observed), color='steelblue4')+
  geom_line(aes(y=detrend), color='darkblue')
```

The renewable energy plot flattened significantly as it lost its upward trend. The hydro plot just shifted down to center at 0, but the shape of the series did not change. This makes sense because the trend of the renewables plot was very influential to the plot's shape, whereas the trend for the hydro series was not.

### Q5

Plot ACF and PACF for the detrended series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. but not mandatory. Did the plots change? How?

```{r}
tre_ts_detrend <- ts(tre_df$detrend, frequency = 12, start(1973,1))

tre_acfplot_dt <- autoplot(Acf(tre_ts_detrend, lag.max=40, plot=F), main=NULL)
                       
tre_pacfplot_dt <- autoplot(Pacf(tre_ts_detrend, lag.max=40,plot=F),main=NULL)

plot_grid(tre_acfplot, tre_acfplot_dt, tre_pacfplot, tre_pacfplot_dt)
```

```{r}
hp_ts_detrend <- ts(hp_df$detrend, frequency = 12, start(1973,1))

hp_acfplot_dt <- autoplot(Acf(hp_ts_detrend, lag.max=40, plot=F), main=NULL)
                       
hp_pacfplot_dt <- autoplot(Pacf(hp_ts_detrend, lag.max=40,plot=F),main=NULL)

plot_grid(hp_acfplot, hp_acfplot_dt, hp_pacfplot, hp_pacfplot_dt)
```

The ACF and PACF plots did not appear to change from the removal of the trend. There were a few more prominent lags for the renewables PACF, with lags 12 and 13 having PACFs further from zero, but there were no notable shifts in these plots.

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6

Just by looking at the time series and the acf plots, do the series seem to have a seasonal trend? No need to run any code to answer your question. Just type in you answer below.

Answer: The hydro plots strongly suggestthat there is a seasonal trend, which makes sense since inflows are dependent on weather. The renewables plots do not have an obvious seasonal trend component just from visual assessment.

### Q7

Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) the two time series. Ask R to print the summary of the regression. Interpret the regression output. From the results which series have a seasonal trend? Do the results match you answer to Q6?

```{r}
dummies <- seasonaldummy(energy_ts[,2])  

tre_seas_means_model <- lm(energy_df[,3]~dummies)
summary(tre_seas_means_model)
```

```{r}
dummies <- seasonaldummy(energy_ts[,3])

hp_seas_means_model <- lm(energy_df[,4]~dummies)
summary(hp_seas_means_model)
```

### Q8

Use the regression coefficients from Q7 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

### Q9

Plot ACF and PACF for the deseason series and compare with the plots from Q1. You may use plot_grid() again to get them side by side. nut mot mandatory. Did the plots change? How?
