---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 4 - Due date 02/12/24"
author: "Jaimie Wargo"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A04_Sp23.Rmd"). Then change "Student Name" on line 4 with your name.

Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai.

R packages needed for this assignment: "xlsx" or "readxl", "ggplot2", "forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.

```{r}
#Load/install required package here
library(readxl)
library(ggplot2)
library(forecast)
library(tseries)
library(Kendall)
library(tidyverse)
library(here)
library(cowplot)
```

## Questions

Consider the same data you used for A3 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. For this assignment you will work only with the column "Total Renewable Energy Production".

```{r}
#Importing data set - using readxl package
raw_data <- read_excel(path=here('Data', 
          'Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx'),
          skip = 12, sheet="Monthly Data",col_names=FALSE) 

#Extract the column names from row 11
read_col_names <- read_excel(path=here('Data', 
          'Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx'),
            skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE)

colnames(raw_data) <- read_col_names

#Data frame
energy_df <- data.frame(raw_data[,c('Month', 'Total Biomass Energy Consumption', 
                         'Total Renewable Energy Consumption', 
                         'Hydroelectric Power Consumption')])

nospace_colnames <- c('date', 'biomass', 'total_renew', 'hydro')
colnames(energy_df) <- nospace_colnames

#Time series structure
energy_ts <- ts(energy_df[,2:4], start=c(1973, 1), frequency = 12)

```

## Stochastic Trend and Stationarity Tests

### Q1

Difference the "Total Renewable Energy Production" series using function diff(). Function diff() is from package base and take three main arguments: *x* vector containing values to be differenced; *lag* integer indicating with lag to use; *differences* integer indicating how many times series should be differenced.

Try differencing at lag 1 only once, i.e., make `lag=1` and `differences=1`. Plot the differenced series Does the series still seem to have trend?

```{r}
energy_df$diff_renewable <- c(diff(energy_df$total_renew, lag=1, differences = 1),0)

ggplot(energy_df, aes(x=date, y=diff_renewable))+
  geom_line(color='deepskyblue4')+
  labs(x='Date', y='Differenced Renewable Series')
```

The series does not appear to have a trend anymore, as the observations are centered on 0 with no increasing or decreasing.

### Q2

Copy and paste part of your code for A3 where you run the regression for Total Renewable Energy Production and subtract that from the original series. This should be the code for Q3 and Q4. make sure you use the same name for you time series object that you had in A3.

```{r}
nobs <- nrow(energy_df)
t <- 1:nobs

renewable_trend <- lm(energy_df$total_renew~t)
summary(renewable_trend)

tre_beta0 <- renewable_trend$coefficients[1]
tre_beta1 <- renewable_trend$coefficients[2]

tre_detrend <- energy_df$total_renew - (tre_beta0+tre_beta1*t)

tre_df <- data.frame('date'=energy_df$date,
                         'observed'=energy_df[,3],
                         'detrend'=tre_detrend)
```

### Q3

Now let's compare the differenced series with the detrended series you calculated on A3. In other words, for the "Total Renewable Energy Production" compare the differenced series from Q1 with the series you detrended in Q2 using linear regression.

Using autoplot() + autolayer() create a plot that shows the three series together. Make sure your plot has a legend. The easiest way to do it is by adding the `series=` argument to each autoplot and autolayer function. Look at the key for A03 for an example.

```{r}
#RenewableEnergyProduction 
ts_plot<-autoplot(energy_ts[,2])+ 
  ylab("Energy [Trillion Btu]")+ 
  ggtitle("") 

diff_plot<-ggplot(energy_df, aes(x=date, y=diff_renewable))+
  geom_line(color='deepskyblue4')+
  labs(x='Date', y='Differenced Series')

detrended_plot<-ggplot(tre_df, aes(x=date, y=detrend))+
  geom_line(color='steelblue3')+
  labs(x='Date', y='LM Detrended Series')

#Addingtitle 
plot_row<-plot_grid(ts_plot,diff_plot,detrended_plot,nrow=1,ncol=3) 

title<-ggdraw()+
  draw_label("Renewable Energy Consumption",fontface='bold') 

plot_grid(title,plot_row,nrow=2,ncol=1,rel_heights=c(0.1,1))

```

### Q4

Plot the ACF for the three series and compare the plots. Add the argument `ylim=c(-0.5,1)` to the autoplot() or Acf() function - whichever you are using to generate the plots - to make sure all three y axis have the same limits. Which method do you think was more efficient in eliminating the trend? The linear regression or differencing?

```{r}
tre_acfplot <- autoplot(Acf(energy_ts[,2], lag.max=40, plot=F),
                        main="Renewable Energy ACF")

tre_ts_detrend <- ts(tre_df$detrend, frequency = 12, start(1973,1))

detrend_acf<- autoplot(Acf(tre_ts_detrend, lag.max=40, plot=F), 
                       main="LInear Detrend")
                       
tre_diff_ts <- ts(energy_df$diff_renewable, frequency = 12, start(1973,1))

diff_acf <- autoplot(Acf(tre_diff_ts, lag.max=40,plot=F),
                     main="Differenced")

plot_row<-plot_grid(tre_acfplot, detrend_acf, diff_acf,nrow=1,ncol=3) 

title<-ggdraw()+
  draw_label("Renewable Energy Consumption ACFs",fontface='bold') 

plot_grid(title,plot_row,nrow=2,ncol=1,rel_heights=c(0.1,1))
```

The differenced series ACF shows that this method was the most efficient in differencing the trend. The values now do not exceed \~0.7. It does seem like there may be some seasonality remaining with strong lags at 12, 24, and 36 months, but overall it succeeded at removing the trend.

### Q5

Compute the Seasonal Mann-Kendall and ADF Test for the original "Total Renewable Energy Production" series. Ask R to print the results. Interpret the results for both test. What is the conclusion from the Seasonal Mann Kendall test? What's the conclusion for the ADF test? Do they match what you observed in Q2? Recall that having a unit root means the series has a stochastic trend. And when a series has stochastic trend we need to use a different procedure to remove the trend.

```{r}


```

### Q6

Aggregate the original "Total Renewable Energy Production" series by year. You can use the same procedure we used in class. Store series in a matrix where rows represent months and columns represent years. And then take the columns mean using function colMeans(). Recall the goal is the remove the seasonal variation from the series to check for trend. Convert the accumulates yearly series into a time series object and plot the series using autoplot().

```{r}

```

### Q7

Apply the Mann Kendal, Spearman correlation rank test and ADF. Are the results from the test in agreement with the test results for the monthly series, i.e., results for Q6?

```{r}

```
